Constrained, yet diverse: Phylogenetic and morphological filtering in epiphytic ferns

### 2019-08-06 ms/manuscript.Rmd
Update conclusion to reflect Testo et al 2018

Here, we show for the first time using a broadly sampled phylogenetic dataset that morphological traits related to epiphytic growth are strongly evolutionarily conserved in ferns. We find that epiphtyic ferns are constrained by size, having consistently smaller fronds and rhizomes than terrerstial species. However, contrary to our hypothesis, epiphytic fern communities appear to be more diverse than terrestrial communities. This suggests that environmental filtering plays a strong role in limiting plant size, while allowing for diversification across other trait axes in epiphytes. Our results are similar between gametophyte and sporophyte traits, suggesting that morphology of each life stage is important for epiphytic growth.

@@

Epiphytes were markedly smaller than terrestrial species as measured by stipe and frond length, frond width, and rhizome diameter at both the species and community level (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)). Stipe length may be a critical trait controlling whole leaf hydraulic conductance; since ferns have resistive stipes in general, species growing in water-limited habitats are expected to decrease stipe length to minimize resistance [@WatkinsJr.2010]. Our results support previous non-phylogenetically controlled studies that found a prevalence of short stipes in epiphytic ferns [@WatkinsJr.2010; @Creese2011]. However, given that these were all correlated with each other (Pearson's correlation coefficient > 0.9), and similar traits have been shown to scale together previously [@Arcand2008; @Creese2011], careful experimental manipulations are needed to distinguish the role (or lack thereof) of each in epiphytic growth.

@@

In contrast to their clearly smaller size, epiphytes were remarkably diverse in other traits. Epiphytes tended to have less dissected fronds with fewer pinna than terrestrial species (\tab \@ref(tab:pics)), but the range of values for these traits was quite wide at the community level (\fig \@ref(fig:cwm-div)). Surprisingly, SLA was only slightly lower in epiphytes relative to terrestrial species (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics). Specific leaf area is an important part of the leaf economic spectrum, which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a]. We hypothesized that epiphytes would have lower SLA, given their more extreme growing conditions. The tradeoffs posited in the leaf economics spectrum have been most thoroughly documented in global studies including the widest possible range of taxa and environments, and do not always hold up in local studies [@Petter2016; @Messier2017]. For example, @Zhu2016 did not find a significant relationship between SLA and leaf life span in a survey of 16 ferns across a disturbance (light) gradient in China. Specific leaf area was the only response variable for which we detected spatial autocorrelation in the residuals (REF SUPP DATA). This, and the strong, parallel relationship in CWM values of SLA with elevation in both epiphytes and terrestrial species (\fig \@ref(fig:cwm-div)), indicates that another unsampled environmental variable may be driving differences in SLA along the gradient.

@@

Results for functional community diversity contradicted our hypotheses; epiphytic communities were more functionally diverse than terrestrial as measured by both MPD and MNTD (\fig \@ref(fig:comm-div)). However, this result is highly dependent on the traits used to measure functional diversity; when the same analyses were done using only size-related traits (scaled, untransformed stipe length, frond length and width, and rhizome diameter), the opposite result was observed (i.e., epiphytic communities are more clustered; \fig S1). Futhermore, terrestrial species show a much wider range of values for size traits (\fig \@ref(fig:cwm-div)), and neither functional group appears to be restricted to a smaller area in trait space (\fig \@ref(fig:pca)). This suggests that epiphytic ferns are subject to extreme environmental filtering for size-related traits, while simultaneously diversifying along other trait axes. This could be maintained by niche partitioning within epiphytic habitats. Epiphytic habitats are highly heterogeneous, and microclimatic conditions vary with height along the host and distance from the trunk [@Hietz1995]. The vertical stratification of epiphytic species into different zones along height gradients in host trees is well documented [@Zotz2007]. In a study comparing terrestrial and epiphytic ferns along an elevational gradient from \SIrange{30}{3000}{\meter} in Costa Rica, @WatkinsJr.2006a found that epiphytes tended to sort into low-trunk and canopy zones, with only 18\% found in both. It seems that the canopy habitat limits overall size in epiphytic ferns, but that species with different traits then sort into different niches. The goal of our study was to quantify differences between epiphytic and terrestrial fern communities; future research should focus on bringing comparative phylogenetic and community ecological methods to bear on the large body of work characterizing differences within epiphytic microhabitats.

@@

We have demonstrated that growth habit, and several traits associated with it, show significant phylogenetic signal in ferns. Furthermore, we find that epiphytes are constrained by size while being remarkably diverse in other traits. Finally, we observed drier conditions in the canopy relative to terrestrial habitats. Taken together, our results support a scenario of environmental filtering acting to limit size and phylogenetic diversity in epiphytic ferns, while simulataneously enabling diversification along other trait axes. The generality of this pattern could be further tested by investigating the functional traits of other groups with large epiphytic radiations such as orchids and bromeliads.

Recent studies [e.g., @Sundue2015; @Testo2016a; @Lehtonen2017] have called into question the scenario of rapid Cretaceous radiations of epiphytic ferns posited by @Schuettpelz2009 by failing to detect increased rates of diversification associated with epiphytic growth. Epiphytism itself is likely a complex trait requiring the evolution of multiple associated traits, and the timing and order of the evolution of such traits may be related to the complex diversification dynamics observed in these studies. Our study identifies several conserved morphological traits (\eg, stipe length, frond dissection, gametophyte morphotype) that may belong to this suite of epiphytic traits. We hope future studies will build on our results by analyzing these traits across much more densely sampled phylogenies [e.g., @Testo2016a] to dissect how each may be related to diversification and growth habit. Additionally, future studies should focus on determining in more detail how niche space is partitioned within epiphytic vs. terrestrial habitats, which is clearly a major dividing line for ferns.

>>>>

Here, we show for the first time using a broadly sampled phylogenetic dataset that multiple morphological traits related to epiphytic growth are evolutionarily conserved in ferns. We find that epiphtyic ferns are constrained by size, having consistently smaller fronds and rhizomes than terrerstial species. However, contrary to our hypothesis, epiphytic fern communities appear to be more diverse than terrestrial communities. This suggests that environmental filtering plays a strong role in limiting plant size, while allowing for diversification across other trait axes in epiphytes.

@@

Epiphytes were markedly smaller than terrestrial species as measured by stipe and frond length, frond width, and rhizome diameter at both the species and community level (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)). Our results at the local scale are corroborated by @Testo2018a, who also observed abrupt size reduction associated with epiphytic growth in a globally sampled fern phylogeny. Stipe length may be a critical trait controlling whole leaf hydraulic conductance; since ferns have resistive stipes in general, species growing in water-limited habitats are expected to decrease stipe length to minimize resistance [@WatkinsJr.2010]. Our results support previous non-phylogenetically controlled studies that found a prevalence of short stipes in epiphytic ferns [@WatkinsJr.2010; @Creese2011]. However, given that these were all correlated with each other (Pearson's correlation coefficient > 0.9), and similar traits have been shown to scale together previously [@Arcand2008; @Creese2011], careful experimental manipulations are needed to distinguish the role (or lack thereof) of each in epiphytic growth.

@@

In contrast to their clearly smaller size, epiphytes were remarkably diverse in other traits. Epiphytes tended to have less dissected fronds with fewer pinna than terrestrial species (\tab \@ref(tab:pics)), but the range of values for these traits was quite wide at the community level (\fig \@ref(fig:cwm-div)). Surprisingly, SLA was only slightly lower in epiphytes relative to terrestrial species (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)). Specific leaf area is an important part of the leaf economic spectrum, which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a]. We hypothesized that epiphytes would have lower SLA, given their more extreme growing conditions. The tradeoffs posited in the leaf economics spectrum have been most thoroughly documented in global studies including the widest possible range of taxa and environments, and do not always hold up in local studies [@Petter2016; @Messier2017]. For example, @Zhu2016 did not find a significant relationship between SLA and leaf life span in a survey of 16 ferns across a disturbance (light) gradient in China. Specific leaf area was the only response variable for which we detected spatial autocorrelation in the residuals (\tab S5). This, and the strong, parallel relationship in CWM values of SLA with elevation in both epiphytes and terrestrial species (\fig \@ref(fig:cwm-div)), indicates that another unsampled environmental variable may be driving differences in SLA along the gradient.

@@

Results for functional community diversity contradicted our hypotheses; epiphytic communities were more functionally diverse than terrestrial as measured by both MPD and MNTD (\fig \@ref(fig:comm-div)). However, this result is highly dependent on the traits used to measure functional diversity; when the same analyses were done using only size-related traits (scaled, untransformed stipe length, frond length and width, and rhizome diameter), the opposite result was observed (\fig S1). Futhermore, terrestrial species show a much wider range of values for size traits (\fig \@ref(fig:cwm-div)), and neither functional group appears to be restricted to a smaller area in trait space (\fig \@ref(fig:pca)). This suggests that epiphytic ferns are subject to extreme environmental filtering for size-related traits, while simultaneously diversifying along other trait axes. This could be maintained by niche partitioning within epiphytic habitats. Epiphytic habitats are highly heterogeneous, and microclimatic conditions vary with height along the host and distance from the trunk [@Hietz1995]. The vertical stratification of epiphytic species into different zones along height gradients in host trees is well documented [@Zotz2007]. In a study comparing terrestrial and epiphytic ferns along an elevational gradient from \SIrange{30}{3000}{\meter} in Costa Rica, @WatkinsJr.2006a found that epiphytes tended to sort into low-trunk and canopy zones, with only 18\% found in both. It seems that the canopy habitat limits overall size in epiphytic ferns, but that species with different traits then sort into different niches. The goal of our study was to quantify differences between epiphytic and terrestrial fern communities; future research should focus on bringing comparative phylogenetic and community ecological methods to bear on the large body of work characterizing differences within epiphytic microhabitats.

@@

We have demonstrated that growth habit, and several traits associated with it, show significant phylogenetic signal in ferns. Furthermore, we find that epiphytes are constrained by size while being remarkably diverse in other traits. Finally, we observed drier conditions in the canopy relative to terrestrial habitats. Taken together, our results support a scenario of environmental filtering acting to limit size and phylogenetic diversity in epiphytic ferns, while simulataneously enabling diversification along other trait axes. This is consistent with a much more broadly sampled study finding that epiphytic and terrestrial ferns do not differ in rates of body size evolution [@Testo2018a]. The generality of this pattern could be further tested by investigating the functional traits of other groups with large epiphytic radiations such as orchids and bromeliads.

Recent studies [e.g., @Sundue2015; @Testo2016a; @Lehtonen2017; @Testo2018a] have called into question the scenario of rapid Cretaceous radiations of epiphytic ferns posited by @Schuettpelz2009 by failing to detect increased rates of diversification associated with epiphytic growth. Epiphytism itself is likely a complex trait requiring the evolution of multiple associated traits, and the timing and order of the evolution of such traits may be related to the complex diversification dynamics observed in these studies. Small size seems to be a prerequisite for epiphytic growth, as indicated both by our study and @Testo2018a. Here, we identify several traits in addition to plant size (\eg, pinna number, frond dissection, gametophyte morphotype) that may also belong to the suite of epiphytic traits. We hope future studies will build on our results by analyzing traits involved in epiphytic growth other than body size across densely sampled phylogenies to dissect how each may relate to diversification and growth habit. Additionally, future studies should focus on determining in more detail how niche space is partitioned within epiphytic vs. terrestrial habitats, which is clearly a major dividing line for ferns.
